---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# nhsbsa

<!-- badges: start -->
[![R-CMD-check](https://github.com/rmgpanw/nhsbsa/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/rmgpanw/nhsbsa/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/rmgpanw/nhsbsa/graph/badge.svg)](https://app.codecov.io/gh/rmgpanw/nhsbsa)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`nhsbsa` is a low-level R client for the [NHS Business Services Authority
(NHSBSA) Open Data Portal](https://opendata.nhsbsa.net), a
[CKAN](https://ckan.org) data catalogue. It provides thin wrappers around the
portal's API actions and returns plain data — tibbles for tabular results and
lists for metadata — leaving the interpretation of any particular dataset to the
caller.

The package deliberately contains no knowledge of specific datasets. Function
names and arguments mirror the CKAN API, so if you know the API you already know
the package.

## Installation

You can install the development version of nhsbsa from
[GitHub](https://github.com/rmgpanw/nhsbsa) with:

```{r}
# install.packages("pak")
pak::pak("rmgpanw/nhsbsa")
```

## Usage

```{r}
library(nhsbsa)
```

Find datasets:

```{r}
# Identifiers of every dataset on the portal
datasets <- nhsbsa_package_list()

# Search for datasets
nhsbsa_package_search(q = "prescribing", rows = 5)
```

List a dataset's resources (files), and download one:

```{r}
resources <- nhsbsa_list_resources(
  "bnf-code-information-current-year",
  pattern = "202401"
)
resources

dest <- nhsbsa_download_resource(
  "bnf-code-information-current-year",
  resource_id = resources$id[[1]]
)
```

Query the rows of a datastore resource without downloading the whole file:

```{r}
# The datastore identifies a resource by its name, e.g. "EPD_202401"
nhsbsa_datastore_search(
  resource_id = "EPD_202401",
  filters = list(pco_code = "13T00"),
  limit = 10
)

# Or with SQL
nhsbsa_datastore_search_sql(
  resource_id = "EPD_202401",
  sql = "SELECT * FROM `EPD_202401` LIMIT 10"
)
```

See `vignette("nhsbsa")` for an overview of the portal and the available
endpoints.
