---
title: "Developer guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Developer guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This guide is for people extending or maintaining `nhsbsa`. It explains how the
package is put together and how to add new functionality consistently. For using
the package, see `vignette("nhsbsa")` instead.

## Design goals

`nhsbsa` is a **low-level, general-purpose client** for the NHS Business Services
Authority Open Data Portal, a [CKAN](https://ckan.org) catalogue. Two principles
shape every decision:

1. **Purity.** There is one exported function per CKAN action, named after it,
   with an argument for every documented API parameter. A user who knows the
   CKAN API should not have to learn anything new. The only deliberate
   exceptions are the two convenience helpers `nhsbsa_list_resources()` and
   `nhsbsa_download_resource()`.
2. **No domain knowledge.** The package knows nothing about specific datasets
   (BNF, prescribing, dental, and so on). It returns plain data — tibbles for
   tabular results, lists for metadata — and leaves interpretation to the
   caller. Dataset-specific logic belongs in a separate package that depends on
   this one.

## Architecture

The package is a thin shell around a single request layer.

```
endpoint functions  ->  nhsbsa_query()  ->  httr2  ->  CKAN
   (R/package.R,         (R/core.R)
    resources.R,
    datastore.R,
    catalogue.R)
```

### The core request layer (`R/core.R`)

`nhsbsa_query(action)` does all the work:

* It inspects the **calling endpoint function** with `rlang::caller_fn()` and
  `rlang::fn_fmls_names()`, reads those argument values from the caller's
  environment, and turns them into query parameters. Arguments prefixed with `.`
  (such as `.return_raw`) are reserved for client behaviour and never sent to
  the API.
* Values are coerced for CKAN by `nhsbsa_format_query_value()`: `NULL`s are
  dropped, logicals become `"true"`/`"false"`, multi-element vectors are joined
  with commas, and lists (such as `filters`) are encoded as JSON.
* The request is performed with retry/backoff. `nhsbsa_perform()` translates
  transport failures into a graceful `nhsbsa_offline` error and HTTP errors into
  `nhsbsa_api_error` (plus a status-specific subclass such as `nhsbsa_http_404`).
* The CKAN envelope is validated: on `success = FALSE` it raises an
  `nhsbsa_api_error`; otherwise it returns the `result` element (or, when the
  caller passed `.return_raw = TRUE`, the whole parsed envelope).

Because parameters are read from the function signature, an endpoint wrapper is
usually a single line — see below.

### Conditions (`R/conditions.R`)

All messages, warnings and errors go through `nhsbsa_abort()`, `nhsbsa_warn()`
and `nhsbsa_inform()`. These wrap the corresponding `cli` functions, always
append a base class (`nhsbsa_error` / `nhsbsa_warning` / `nhsbsa_message`), and
accept an optional `class` to prepend a more specific subclass. Give a condition
a specific subclass whenever a caller might reasonably want to catch it, e.g.

```{r}
nhsbsa_abort(
  c("x" = "No resource matched."),
  class = "nhsbsa_resource_not_found"
)
```

Do not call `cli::cli_abort()` (or any other package's error helpers) directly.

### Return-type conventions

* Catalogue/metadata actions return a **list** (`package_show`, `package_search`,
  `resource_show`).
* Listing actions return a **character vector** (`package_list`,
  `organization_list`, `group_list`, `tag_list`).
* Row queries return a **tibble** (`datastore_search`, `datastore_search_sql`,
  `list_resources`).
* `nhsbsa_download_resource()` returns the destination path invisibly.
* Any read-only function called with `.return_raw = TRUE` returns the parsed
  envelope as a list.

## Adding a new endpoint

To wrap another CKAN action, add a thin function whose arguments mirror the API
parameters. For example, to wrap `tag_show`:

```{r}
#' Show a tag
#'
#' Wraps the CKAN `tag_show` action.
#'
#' @param id Character scalar. The tag name or id.
#' @inheritParams nhsbsa_package_list
#'
#' @return A list of tag metadata.
#' @export
#' @examplesIf identical(Sys.getenv("IN_PKGDOWN"), "true")
#' nhsbsa_tag_show("prescribing")
nhsbsa_tag_show <- function(id, .return_raw = FALSE) {
  nhsbsa_query("tag_show")
}
```

That is the whole implementation: `nhsbsa_query()` picks up `id` from the
signature and sends it as a query parameter. If the action returns tabular data,
post-process the result with `nhsbsa_records_to_tibble()`.

Then:

```{r}
devtools::document() # regenerate NAMESPACE and the .Rd file
```

Keep examples gated with `@examplesIf identical(Sys.getenv("IN_PKGDOWN"), "true")`
so they render on the documentation site but never run on CRAN.

## Testing

Tests use `testthat` (edition 3). The suite is designed to run fully offline and
deterministically; a small set of live tests is gated separately.

### Offline tests with recorded fixtures

Network calls are recorded as [httptest2](https://enpiar.com/httptest2/)
fixtures and replayed. Wrap a block in `with_mock_dir()`; the first run records
the responses, later runs replay them:

```{r}
with_mock_dir("pl", {
  test_that("nhsbsa_package_list returns dataset ids", {
    out <- nhsbsa_package_list()
    expect_type(out, "character")
  })
})
```

Keep the mock directory name **short**. The portal's URL path
(`opendata.nhsbsa.net/api/3/action/`) is deep, and R CMD check rejects tarball
paths longer than 100 bytes, so a long directory name plus a long action name
can tip a fixture over the limit.

### Offline tests with mocked responses

For error paths and edge cases it is simpler to mock the HTTP response directly,
without a fixture, using `httr2::with_mocked_responses()`:

```{r}
test_that("an HTTP error surfaces as nhsbsa_api_error", {
  resp <- httr2::response(
    status_code = 404L,
    headers = list(`Content-Type` = "application/json"),
    body = charToRaw('{"success":false,"error":{"message":"nope"}}')
  )
  httr2::with_mocked_responses(function(req) resp, {
    expect_error(nhsbsa_package_show("x"), class = "nhsbsa_http_404")
  })
})
```

### Live tests

Tests that contact the real API live in `tests/testthat/test-live.R`. They are
skipped on CRAN, skipped when offline, and only run when the `NHSBSA_LIVE_TESTS`
environment variable is set. The continuous integration workflows set this
variable so the live tests run there. Use live tests sparingly, as a smoke test
that the real API still behaves as expected.

## Checks before a pull request

```sh
air format .
```

```r
devtools::document()
devtools::test()
lintr::lint_package()
spelling::spell_check_package()
devtools::check()
```

Aim for a clean `devtools::check()` (no errors, warnings or notes). New
user-facing terms that the spell checker flags can be added to
`inst/WORDLIST`.
