---
title: "Getting started with nhsbsa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with nhsbsa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(nhsbsa)
```

## The NHSBSA Open Data Portal

The [NHS Business Services Authority (NHSBSA) Open Data
Portal](https://opendata.nhsbsa.net) publishes open datasets about NHS
activity in England — prescribing, dental, pharmaceutical and contractor data
among them. All of it is freely available under the Open Government Licence.

The portal runs on [CKAN](https://ckan.org), a widely used open-source data
catalogue. CKAN organises data into **datasets** (called "packages") which each
contain one or more **resources** — the individual downloadable files (usually
CSV) or database tables behind them.

`nhsbsa` is a thin, low-level client for this portal. It wraps the CKAN API and
returns plain data for you to work with: tibbles for tabular results and lists
for metadata. It contains no knowledge of any particular dataset, so you supply
the dataset identifiers and interpret the results yourself. If you are familiar
with the CKAN API, the package will feel familiar too: function names and
arguments mirror the API.

For background on the underlying API, see the portal's own
[API page](https://opendata.nhsbsa.net/pages/api) and the
[CKAN Action API reference](https://docs.ckan.org/en/latest/api/).

## The API and the response envelope

Every request goes to a CKAN *action* under

```
https://opendata.nhsbsa.net/api/3/action/<action>
```

and comes back as a JSON envelope of the form

```json
{ "success": true, "result": ... }
```

or, on failure,

```json
{ "success": false, "error": { "message": "..." } }
```

`nhsbsa` handles this envelope for you: it checks the `success` flag, returns
the `result`, and otherwise raises an informative error. If the portal cannot be
reached (for example with no internet connection) it fails gracefully with a
clear message rather than an obscure low-level error.

## What the package wraps

| Function | CKAN action | Returns |
|---|---|---|
| `nhsbsa_package_list()` | `package_list` | character vector of dataset ids |
| `nhsbsa_package_show()` | `package_show` | list (dataset metadata) |
| `nhsbsa_package_search()` | `package_search` | list (search results) |
| `nhsbsa_resource_show()` | `resource_show` | list (resource metadata) |
| `nhsbsa_datastore_search()` | `datastore_search` | tibble of rows |
| `nhsbsa_datastore_search_sql()` | `datastore_search_sql` | tibble of rows |
| `nhsbsa_organization_list()` | `organization_list` | character vector of organisations |
| `nhsbsa_group_list()` | `group_list` | character vector of groups |
| `nhsbsa_tag_list()` | `tag_list` | character vector of tags |

Two convenience helpers sit on top of these for the operations you reach for
most often. They are not pure API wrappers, but they save a little work:

* `nhsbsa_list_resources()` flattens a dataset's resources into a tibble
  (including each resource's download URL).
* `nhsbsa_download_resource()` resolves a single resource and streams its file
  to disk.

Most read-only functions also accept `.return_raw = TRUE`, which returns the
full parsed response envelope instead of the processed result — useful when you
need fields the helper does not surface, such as a datastore query's `total`.

## From the portal website to the API

It helps to think of the package as a programmatic version of the
[portal website](https://opendata.nhsbsa.net). The things you click there map
onto API calls:

* **Browsing all datasets** is `nhsbsa_package_list()` (identifiers) or
  `nhsbsa_package_search()` (richer records, with searching and paging).
* **Clicking a tag**, say **#Prescribing**, takes the website to
  `/dataset/?tags=Prescribing`. The equivalent call uses a *filter query* on the
  `tags` field:

    ```{r}
    nhsbsa_package_search(fq = 'tags:"Prescribing"')
    ```

  Tags are case-sensitive; list them with `nhsbsa_tag_list()`. Similarly,
  filtering by **organisation** maps to `fq = 'organization:<name>'`, with names
  from `nhsbsa_organization_list()`.
* **Opening a dataset's page** (e.g.
  `/dataset/english-prescribing-data-epd`) corresponds to
  `nhsbsa_package_show("english-prescribing-data-epd")`, whose `resources`
  element lists the files shown on that page. `nhsbsa_list_resources()` tidies
  those resources into a tibble.
* **The "Download" button** on a resource fetches the file that
  `nhsbsa_download_resource()` streams to disk; **the data preview / "Data API"**
  for a resource is backed by the datastore that `nhsbsa_datastore_search()`
  queries.

## A worked example

### List datasets

```{r}
datasets <- nhsbsa_package_list()
length(datasets)
head(datasets)
```

### Inspect a dataset's resources

```{r}
resources <- nhsbsa_list_resources("bnf-code-information-current-year")
resources
```

You can filter by a pattern matched against the resource name:

```{r}
nhsbsa_list_resources(
  "bnf-code-information-current-year",
  pattern = "202401"
)
```

### Download a resource file

Identify a single resource — by `resource_id`, or by a `pattern` that matches
exactly one resource name — and stream it to disk:

```{r}
dest <- nhsbsa_download_resource(
  "bnf-code-information-current-year",
  pattern = "202401"
)
dest
```

### Query rows without downloading the whole file

When a resource is loaded into the datastore you can query its rows directly,
without downloading the whole file. Note that the datastore identifies a
resource by its **name** (e.g. `"EPD_202401"`), shown in the `name` column of
`nhsbsa_list_resources()`, rather than by its `id`.

## Ways to query datastore data

There are three ways to narrow a datastore query, from simplest to most
powerful. Use whichever fits the question.

### 1. `filters` — exact field matching

`filters` takes a named list of field/value pairs and returns rows where each
field equals the given value exactly. This is the most reliable option. Field
names are case-sensitive and must match the resource's columns exactly (for the
EPD they are upper case, e.g. `PCO_CODE`):

```{r}
nhsbsa_datastore_search(
  resource_id = "EPD_202401",
  filters = list(PCO_CODE = "13T00", BNF_CHEMICAL_SUBSTANCE = "0407010H0"),
  limit = 10
)
```

The datastore returns at most one page of rows per request. If more rows match
than were returned, `nhsbsa_datastore_search()` warns you and explains how to
page through the rest by increasing `offset`:

```{r}
# Fetch the next page
nhsbsa_datastore_search(
  resource_id = "EPD_202401",
  filters = list(PCO_CODE = "13T00"),
  limit = 10,
  offset = 10
)
```

### 2. `q` — full-text search

For a *per-field* full-text search, pass `q` a named list. Each value is matched
as text within that field (rather than required to be exactly equal):

```{r}
nhsbsa_datastore_search(
  resource_id = "EPD_202401",
  q = list(BNF_CHEMICAL_SUBSTANCE = "Paracetamol"),
  limit = 10
)
```

As a plain string, `q` searches across all fields:

```{r}
nhsbsa_datastore_search(resource_id = "EPD_202401", q = "Paracetamol")
```

Note that full-text search is not enabled for every resource on the portal — a
plain-string `q` may return a server error for some resources, in which case use
`filters` or SQL instead.

### 3. SQL — full query language

For column selection, expressions, aggregation, joins and sorting, use SQL. The
portal requires the `resource_id` alongside the query, and you reference the same
resource name in the `FROM` clause:

```{r}
nhsbsa_datastore_search_sql(
  resource_id = "EPD_202401",
  sql = "SELECT PCO_CODE, SUM(ITEMS) AS items
         FROM `EPD_202401`
         GROUP BY PCO_CODE
         ORDER BY items DESC"
)
```

The SQL string is sent to the API verbatim, so you are responsible for paging
(via `LIMIT`/`OFFSET`) and for quoting identifiers correctly.
