---
title: "Getting started with nhsbsa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with nhsbsa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(nhsbsa)
```

## The NHSBSA Open Data Portal

The [NHS Business Services Authority (NHSBSA) Open Data
Portal](https://opendata.nhsbsa.net) publishes open datasets about NHS
activity in England — prescribing, dental, pharmaceutical and contractor data
among them. All of it is freely available under the Open Government Licence.

The portal runs on [CKAN](https://ckan.org), a widely used open-source data
catalogue. CKAN organises data into **datasets** (called "packages") which each
contain one or more **resources** — the individual downloadable files (usually
CSV) or database tables behind them.

`nhsbsa` is a thin, low-level client for this portal. It wraps the CKAN API and
returns plain data for you to work with: tibbles for tabular results and lists
for metadata. It contains no knowledge of any particular dataset, so you supply
the dataset identifiers and interpret the results yourself. If you are familiar
with the CKAN API, the package will feel familiar too: function names and
arguments mirror the API.

## The API and the response envelope

Every request goes to a CKAN *action* under

```
https://opendata.nhsbsa.net/api/3/action/<action>
```

and comes back as a JSON envelope of the form

```json
{ "success": true, "result": ... }
```

or, on failure,

```json
{ "success": false, "error": { "message": "..." } }
```

`nhsbsa` handles this envelope for you: it checks the `success` flag, returns
the `result`, and otherwise raises an informative error. If the portal cannot be
reached (for example with no internet connection) it fails gracefully with a
clear message rather than an obscure low-level error.

## What the package wraps

| Function | CKAN action | Returns |
|---|---|---|
| `nhsbsa_package_list()` | `package_list` | character vector of dataset ids |
| `nhsbsa_package_show()` | `package_show` | list (dataset metadata) |
| `nhsbsa_package_search()` | `package_search` | list (search results) |
| `nhsbsa_resource_show()` | `resource_show` | list (resource metadata) |
| `nhsbsa_datastore_search()` | `datastore_search` | tibble of rows |
| `nhsbsa_datastore_search_sql()` | `datastore_search_sql` | tibble of rows |

Two convenience helpers sit on top of these for the operations you reach for
most often. They are not pure API wrappers, but they save a little work:

* `nhsbsa_list_resources()` flattens a dataset's resources into a tibble
  (including each resource's download URL).
* `nhsbsa_download_resource()` resolves a single resource and streams its file
  to disk.

Most read-only functions also accept `.return_raw = TRUE`, which returns the
full parsed response envelope instead of the processed result — useful when you
need fields the helper does not surface, such as a datastore query's `total`.

## A worked example

### List datasets

```{r}
datasets <- nhsbsa_package_list()
length(datasets)
head(datasets)
```

### Inspect a dataset's resources

```{r}
resources <- nhsbsa_list_resources("bnf-code-information-current-year")
resources
```

You can filter by a pattern matched against the resource name:

```{r}
nhsbsa_list_resources(
  "bnf-code-information-current-year",
  pattern = "202401"
)
```

### Download a resource file

Identify a single resource — by `resource_id`, or by a `pattern` that matches
exactly one resource name — and stream it to disk:

```{r}
dest <- nhsbsa_download_resource(
  "bnf-code-information-current-year",
  pattern = "202401"
)
dest
```

### Query rows without downloading the whole file

When a resource is loaded into the datastore you can query its rows directly.
Note that the datastore identifies a resource by its **name** (e.g.
`"EPD_202401"`), shown in the `name` column of `nhsbsa_list_resources()`, rather
than by its `id`:

```{r}
nhsbsa_datastore_search(
  resource_id = "EPD_202401",
  filters = list(pco_code = "13T00"),
  limit = 10
)
```

The datastore returns at most one page of rows per request. If more rows match
than were returned, `nhsbsa_datastore_search()` warns you and explains how to
page through the rest by increasing `offset`:

```{r}
# Fetch the next page
nhsbsa_datastore_search(
  resource_id = "EPD_202401",
  filters = list(pco_code = "13T00"),
  limit = 10,
  offset = 10
)
```

For more control you can query with SQL. The portal requires the `resource_id`
alongside the query, and you reference the same resource name in the `FROM`
clause:

```{r}
nhsbsa_datastore_search_sql(
  resource_id = "EPD_202401",
  sql = "SELECT pco_code, SUM(items) AS items
         FROM `EPD_202401`
         GROUP BY pco_code"
)
```
